TBD

### 2019-09-18 manuscript/manuscript.tex
updated signs of selection gradients in Table 2
update manuscript with Chris' suggestions and attempt moving figures to fit nicely.

\documentclass[11pt,]{article}
\usepackage[]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=black,
            urlcolor=black,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{center} \LARGE Consumer extinctions constrain phenotypic evolution in the resulting food web \end{center}

@@

\bigskip
<!----> \emph{Manuscript elements}: table 1, table 2, figure 1, figure
2, figure 3, figure 4. All figures should be printed in color.

@@

\emph{Keywords}: adaptive landscape, ecological networks,
eco-evolutionary dynamics, natural selection, host-parasitoid;
complexity.

@@

\emph{Number of words}: 3109

@@

\emph{Manuscript type}: Article.

@@

\footnotesize Prepared using an \emph{Am. Nat.} inspired \LaTeX{}
template for Rmarkdown. \normalsize

@@

Global change is altering the structure of ecological networks; however,
we are currently in a poor position to predict how these altered
communities will affect the evolutionary potential of remaining
populations. Theory on adaptive landscapes provides a framework for
predicting how selection constrains phenotypic evolution, but often
treats the community context of evolving populations as a ``black box''.
Here, we integrate ecological networks and adaptive landscapes to
examine how changes in food-web structure shape evolutionary
constraints. We conducted a field experiment that simulated the
extinction of a guild of larval parasitoids that were able to impose
selection on an insect herbivore. We then measured herbivore survival as
a function of three key phenotypic traits. We found that the number of
traits under selection increased with the extinction of larval
parasitoids. In contrast, the adaptive landscape was more neutral in the
original food web because different parasitoid guilds impose different
selection pressures, minimizing relative fitness differences among
phenotypes. Our results suggest that the loss of trophic interactions
due to consumer extinctions can impose greater constraints on phenotypic
evolution. This indicates that the simplification of ecological
communities may constrain the adaptive potential of remaining
populations to future environmental change.

@@

The adaptive landscape provides a powerful framework for understanding
how natural selection has shaped the evolution of biodiversity ---from
genes to phenotypes to species
\citep{Wright1931, Simpson1944, Arnold2001}. More than a metaphor, the
adaptive landscape links quantitative genetic and phenotypic variation
to evolution by natural selection
\citep{Lande1979, Arnold1984applications, Arnold1984theory}. Ecological
interactions often play a key role in shaping adaptive landscapes, as
evidenced by the role of antagonistic \citep{Schluter2000, Abrams2000}
and mutualistic \citep{Bronstein2006} interactions in driving
evolutionary change. Although there is clear evidence that pairwise
interactions can shape the adaptive landscape, we still have a poor
understanding of how the adaptive landscape is shaped by the community
context \citep{McPeek2017, terHorst2018}. Resolution on how the
community context shapes phenotypic evolution is urgently needed though,
given the rapid impacts of climate change on ecological communities
\citep{Scheffers2016}.

Ecological networks, such as food webs describing who-eats-whom in
ecological communities, provide an explicit representation of the
community context \citep{Bascompte2014, McCann2012}. Here, we integrate
ecological networks and adaptive landscapes to understand how the
community context constrains evolutionary change \citep{Arnold1992}.
Different aspects of evolutionary constraints can be inferred by
quantifying the slope and curvature of the adaptive landscape. For
example, the slope is determined by directional selection gradients
acting on each phenotypic trait and influences the trajectory of
evolutionary change \citep{Lande1979, Arnold1992}. Selective constraints
on evolution increase with the number of traits under directional
selection, as this diminishes the number of optimal solutions
\citep{Arnold2003}. The curvature of the adaptive landscape can also
constrain evolution through its indirect effect on genetic constraints
\citep{Arnold1992, Hansen2008}. Genetic constraints are largely governed
by a population's \textbf{G}-matrix ---the additive genetic variances
and covariances between traits \citep{Hansen2008}. In general, genetic
constraints will increase with the number of traits under directional or
stabilizing selection, as this will decrease the additive genetic
variance in those traits \citep{Hansen2008}. Genetic constraints may
also increase with the number of trait combinations under correlational
selection, as this type of selection decreases the evolutionary
independence of traits \citep{Hansen2008}. If we want to predict how the
community context constrains evolutionary change, we must understand how
ecological networks shape the adaptive landscape.

Global change is altering the structure of ecological networks, which
may influence evolutionary constraints in a number of ways. For example,
in a food web, if different consumers impose directional selection on
different traits of a shared resource, then a more diverse consumer
community may constrain evolution by increasing the number of traits
under selection. Alternatively, if consumers impose selection on
different values of a trait, then their selective effects would cancel
each other out in more diverse communities. To examine these different
possibilities, we conducted a field experiment that simulated the
extinction of a consumer guild associated with an abundant insect
herbivore (\emph{Iteomyia salicisverruca}, Family Cecidomyiidae; fig.
\ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped
galls when they feed on the developing leaves of willow trees
\citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva from
attack by generalist predators (e.g.~ants, spiders), but they suffer
high mortality from egg and larval parasitoids \citep{Barbour2016}. We
manipulated food-web structure by either excluding or allowing the guild
of larval parasitoids to impose selection on gall midge and egg
parasitoid traits (fig. \ref{fig:Conceptual}). We then applied modern
statistical methods to quantify how changes in food-web structure
altered the slope and curvature of the adaptive landscape of the gall
midge and egg parasitoid. Taken together, our study gives insight to how
local extinctions may constrain the evolution of interacting
populations.

\begin{figure}
\centering
\includegraphics{analyses/complex_simple_foodwebs_v2.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. To simulate consumer
extinction, we used mesh bags to exclude the guild of larval
parasitoids, only allowing the egg parasitoid (\emph{Platygaster} sp.)
to impose selection. Larval parasitoids include the following species
(from left to right): \emph{Mesopolobus} sp. (Family: Pteromalidae);
\emph{Tetrastichus} sp. (Family: Eulophidae); and \emph{Torymus} sp.
(Family: Torymidae).}
\end{figure}

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from eight different plant genotypes that spanned the range of trait
variation observed in this willow population \citep{Barbour2015}. For
the original food web (eight replicates per genotype), we used flagging
tape to mark 14 galled leaves per plant (\textasciitilde{}30 larvae),
allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that all galls had
similar phenology when we collected galls later in the season. To
simulate consumer extinctions, we enclosed 14 galled leaves with 10x15cm
organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three
parasitoid species that attack during larval development. This treatment
did not exclude the egg parasitoid \emph{Platygaster} sp., which attacks
prior to gall initiation \citep[larva initiate gall development in
Cecidomyiid midges:][]{Gagne1989}. In late August, we collected marked
and bagged galls from each plant, placed them into 30 mL vials and kept
them in the lab for 4 months at room temperature. We then opened galls
under a dissecting scope and determined whether larvae survived to
pupation (our measure of fitness) or were parasitized. We did not
include other sources of mortality, such as early larval death, as they
could influence the expression of the gall phenotype and confound
estimates of selection. For the food-web treatment that excluded larval
parasitoids, we further restricted our data by removing any incidental
instances of parasitism by a larval parasitoid. This represented less
than 3\% of the observations in this food-web treatment and allowed us
to focus our inferences of selection on those imposed by the egg
parasitoid. Our final dataset contains survival estimates for 1285
larvae from 613 galls and 111 plants.

@@

Inferring adaptive landscapes assumes that trait distributions are
multivariate normal \citep{Lande1983}. To approximate this assumption,
we log-transformed clutch size and square-root transformed oviposition
preference. We then scaled all phenotypic traits (mean=0 and SD=1) prior
to our analyses (detailed below) to ensure that our estimates of
selection were comparable across traits and with other studies.

Our analysis consisted of three parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
standardized phenotypic traits (\(i\)) of individuals \textemdash in
each food-web treatment. Second, we translated selection surfaces into
the scale of relative fitness (\(w\)) in order to calculate standardized
selection gradients. Third, we used our estimates of selection gradients
to characterize the slope and curvature of the adaptive landscape, which
we used to measure evolutionary constraints.

\textbf{Selection surfaces}: Since larval survival was our measure of
absolute fitness, we used a GLMM that assumed a binomial error
distribution (and logit-link function). To approximate the selection
surface, we modelled larval survival as a function of food-web treatment
as well as linear (\(\alpha_i\)), quadratic (\(\alpha_{ii}\)), and
linear interactions (\(\alpha_{ij}\)) between each trait. We also
allowed these selection surfaces (\(\alpha\)) to vary between food-web
treatments. Note that to obtain valid estimates of linear selection
surfaces, we removed nonlinear terms prior to estimating linear
relationships \citep{Lande1983}. Other approaches have been advocated
for approximating selection surfaces \citep{Schluter1988}; however, our
approach enables us to calculate selection gradients, and thus is more
appropriate for approximating the adaptive landscape \citep{Arnold2003}.
To account for the nonindependence of clutch size (measured at gall
level) and oviposition preference (measured at plant level) as well as
any independent effects of willow genotype on larval survival, we
modelled gall ID nested within plant ID nested within genotype ID as
random effects. Although statistical models with random effects are not
common in analyses of natural selection, we think that modelling random
effects can mitigate biased estimates of selection due to environmental
covariances between traits and fitness \citep{Rausher1992}. Since our
end goal was to characterize the relationship between mean trait values
and mean fitness (adaptive landscape), we assumed the mean value of our
random effects (i.e., setting them to zero) when calculating selection
surfaces. We then used parametric bootstrapping (1,000 replicates) to
estimate the effect of food-web treatment on larval survival as well as
selection surfaces in each food-web treatment.

\textbf{Selection gradients}: Selection gradients cannot be estimated
directly from GLMMs of selection surfaces because the response is in
terms of absolute fitness and the coefficients are on a nonlinear scale.
For example, the coefficients in the above model measure the change in
the logarithm of the odds of surviving (i.e.,
\(\text{ln}\{W(z)/[1-W(z)]\}\)) associated with 1 SD change in a trait
with all other traits held fixed at their mean. Therefore, we used the
method developed by \citet{Janzen1998} to translate selection surfaces
from the above model into the scale of relative fitness in order to
calculate directional (\(\beta_i\)), quadratic (\(\gamma_{ii}\)), and
correlational (\(\gamma_{ij}\)) selection gradients. Briefly, this
method calculates the average gradient of selection surfaces by
multiplying the average of \(W(z)[1-W(z)]\) by each regression
coefficient (e.g. \(\alpha_i\), \(\alpha_{ii}\), or \(\alpha_{ij}\)). We
then divided this average gradient by the mean fitness (\(\bar W\)) to
put it on the scale of relative fitness, and thus interpretable as a
selection gradient. We estimated selection gradients separately for each
food-web treatment. Note that we doubled all quadratic terms prior to
calculating selection gradients to put them on the same scale as
estimates of directional and correlational selection
\citep{Stinchcombe2008}.

\textbf{Evolutionary constraints}: We measured evolutionary constraints
by inspecting the slope and curvature of the adaptive landscape. The
number of selective constraints is determined by the slope of the
adaptive landscape, which in our study corresponds to:

\[\textbf{Slope} = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} \]
where each \(\beta_i\) corresponds to the directional selection gradient
acting on each trait. By comparing the number of directional selection
gradients that show clear evidence of contributing to the slope (i.e.,
95\% CI does not overlap zero) in our food-web treatments, we can infer
the effect of food-web structure on selective constraints.

The indirect effects of selection on the G-matrix
(\(\Delta\text{G}=\text{G}(\gamma - \beta \beta^\text{T})\text{G}\)) is
governed by the curvature of the adaptive landscape
(\(\text{C}=\gamma - \beta \beta^\text{T}\)), which in our study
corresponds to:

@@

\[\textbf{Curvature} = \begin{pmatrix} \text{C}_{\text{Diam:Diam}}&& \\ \text{C}_{\text{Clutch:Diam}} & \text{C}_{\text{Clutch:Clutch}} & \\ \text{C}_{\text{Pref:Diam}} & \text{C}_{\text{Pref:Clutch}} & \text{C}_{\text{Pref:Pref}} \end{pmatrix}\]
where each \(\text{C}_{ii}\) (diagonal) represents the effect of
selection on the additive genetic variance in a trait, and each
\(\text{C}_{ij}\) (off-diagonal) represents the effect of selection on
the additive genetic covariance between a particular trait combination.
In other words, the sign of diagonal terms of the curvature matrix
dictate whether selection will increase (+), decrease (-), or cause no
change (0) in the additive genetic variance of a trait. Similarly, any
nonzero covariance terms (off-diagonal) are indicative of selection for
trait integration (i.e., less independence). Therefore, we can infer the
number of genetic constraints by summing the number of negative signs
along the diagonal (decrease in additive genetic variance) and the
number of nonzero terms along the off-diagonal (trait integration) of
the curvature matrix. For these analyses, we retained estimates for
selection gradients that showed clear evidence of being different from
zero (i.e., 95\% CI did not overlap zero) and set other gradients to
zero, prior to calculating the curvature matrix in each food-web
treatment. This provides a conservative estimate of the factors
contributing to the curvature of the adaptive landscape.

>>>>

\documentclass[11pt,]{article}
\usepackage[]{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            colorlinks=true,
            linkcolor=black,
            citecolor=black,
            urlcolor=black,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{center} \LARGE Loss of consumers constrains phenotypic evolution in the resulting food web \end{center}

@@

@@

\emph{Short Running Title}: Consumer loss constrains phenotypic
evolution

@@

\emph{Keywords}: adaptive landscape, ecological networks,
eco-evolutionary dynamics, natural selection, host-parasitoid,
complexity, biodiversity, community context.

@@

\emph{Abstract Word Count}: 251

@@

\emph{Total Word Count}: 5769

\emph{Manuscript type}: Letter.

@@

The loss of biodiversity is altering the structure of ecological
networks; however, we are currently in a poor position to predict how
these altered communities will affect the evolutionary potential of
remaining populations. Theory on adaptive landscapes provides a
framework for predicting how selection constrains phenotypic evolution,
but often treats the network context of evolving populations as a
``black box''. Here, we integrate ecological networks and adaptive
landscapes to examine how changes in food-web structure shape
evolutionary constraints. We conducted a field experiment that removed a
guild of larval parasitoids that imposed direct and indirect selection
pressures on an insect herbivore. We then measured herbivore survival as
a function of three key phenotypic traits to estimate directional,
quadratic, and correlational selection gradients in each treatment. We
used these selection gradients to characterize the slope and curvature
of the adaptive landscape to understand the direct and indirect effects
of consumer loss on evolutionary constraints. We found that the number
of traits under directional selection increased with the removal of
larval parasitoids, indicating greater selective constraints on the
trajectory of evolutionary change. Similarly, we found that the removal
of larval parasitoids altered the curvature of the adaptive landscape in
such a way that tended to decrease the evolvability of the traits we
measured in the next generation. Our results suggest that the loss of
trophic interactions can impose greater constraints on phenotypic
evolution. This indicates that the simplification of ecological
communities may constrain the adaptive potential of remaining
populations to future environmental change.

\section{Impact Summary}\label{impact-summary}

The loss of biodiversity is rewiring the web of life; however, it is
uncertain how this will affect the ability of remaining populations to
evolve and adapt to future environments. To gain insight to these
effects, we conducted a field experiment that either maintained a
natural community of predators or removed all but one of the predators
that was able to impose selection on a common prey. We found that the
loss of predators acted to constrain prey evolution toward a particular
combination of traits. Moreover, we found that the loss of predators
could make it more difficult for prey to adapt to uncertain future
environments. Taken together, our results suggest that the
simplification of the web of life may constrain the adaptive potential
of remaining populations.

@@

The adaptive landscape provides a powerful framework for understanding
how natural selection has shaped the evolution of biodiversity ---from
genes to phenotypes to species
\citep{Wright1931, Simpson1944, Arnold2001}. More than a metaphor, the
adaptive landscape links quantitative genetic and phenotypic variation
to evolution by natural selection
\citep{Lande1979, Arnold1984applications, Arnold1984theory}, giving
insight to the direct and indirect effects of selective constraints
\citep{Arnold1992}. Ecological interactions often play a key role in
shaping selective constraints, as evidenced by the role of antagonistic
and mutualistic interactions in driving evolutionary change
\citep{Schluter2000, Abrams2000, Bronstein2006}. Although there is clear
evidence that species interactions can shape the adaptive landscape, we
still have a poor understanding of how the adaptive landscape is shaped
by a community of interacting species \citep{McPeek2017, terHorst2018}.
Resolution on how change in ecological communities shape phenotypic
evolution is urgently needed though, given the rapid losses of
biodiversity we are observing in the Anthropocene \citep{Scheffers2016}.

Ecological networks, such as food webs describing who-eats-whom, provide
an explicit representation of the direct and indirect effects that
emerge in a community of interacting species
\citep{Bascompte2014, McCann2012}. Here, we integrate ecological
networks and adaptive landscapes to understand how ecological
communities constrain evolutionary change. The direct and indirect
effects of natural selection can be inferred by quantifying the slope
and curvature of the adaptive landscape \citep{Arnold1992}. For example,
the slope is determined by directional selection gradients acting on
each phenotypic trait and influences the trajectory of evolutionary
change \citep{Lande1979, Arnold1992}. Selective constraints on evolution
increase with the number of traits under directional selection, as this
diminishes the number of optimal solutions \citep{Arnold2003}. The
curvature of the adaptive landscape is governed by nonlinear selection
gradients and acts to indirectly constrain evolution through its effect
on genetic constraints \citep{Arnold1992, Hansen2008}. Genetic
constraints are largely governed by a population's \textbf{G}-matrix
---the additive genetic variances and covariances between traits
\citep{Hansen2008}. Stabilizing selection acts to erode genetic variance
in traits, which can impose a constraint on the ability of those traits
to respond to future selection \citep{Hansen2008}. Correlational
selection may alter the genetic covariance between traits, which may
facilitate or hinder future adaptation depending on pattern of selection
on those traits \citep{Hansen2008}. If we want to predict how ecological
communities shape phenotypic evolution, we must understand how
ecological networks shape the adaptive landscape.

The loss of biodiversity is altering the structure of ecological
networks, which may influence evolutionary constraints in a number of
ways. For example, in a food web, if different consumers impose
directional selection on different traits of a shared resource, then a
more diverse consumer community may constrain evolution by increasing
the number of traits under selection. Alternatively, if consumers impose
selection on different values of a trait, then their selective effects
would cancel each other out in more diverse communities. To examine
these different possibilities, we conducted a field experiment that
removed a consumer guild that directly and indirectly impacts fitness of
an abundant insect herbivore (\emph{Iteomyia salicisverruca}, Family
Cecidomyiidae; fig. \ref{fig:Conceptual}). The larvae of this herbivore
induce tooth-shaped galls when they feed on the developing leaves of
willow trees \citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect
larva from attack by generalist predators (e.g.~ants, spiders), but they
suffer high mortality from egg and larval parasitoids
\citep{Barbour2016}. We manipulated food-web structure by either
removing larval parasitoids (removal food web) or allowing both egg and
larval parasitoids to impose selection on gall midge traits (original
food web, fig. \ref{fig:Conceptual}). The relative abundance of these
larval parasitoids is much lower than the egg parasitoid
\citep{Barbour2016}, and thus represents a plausible extinction scenario
for this system. Larval parasitoids also impose indirect effects on gall
midge fitness through intraguild predation on the egg parasitoid (fig.
\ref{fig:Conceptual}). We applied modern statistical methods to quantify
how changes in food-web structure altered the slope and curvature of the
gall midge's adaptive landscape. Taken together, our study gives insight
to how the loss of biodiversity may constrain the evolution of
interacting populations.

\begin{figure}
\centering
\includegraphics{../analyses/complex_simple_foodwebs_v3.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. In the removal food web, we
used mesh bags to exclude the guild of larval parasitoids, only allowing
the egg parasitoid (\emph{Platygaster} sp.) to impose selection. Note
that larval parasitoids also impose indirect effects on gall midge
fitness through intraguild predation on the egg parasitoid. Larval
parasitoids include the following species (from left to right):
\emph{Mesopolobus} sp. (Family: Pteromalidae); \emph{Tetrastichus} sp.
(Family: Eulophidae); and \emph{Torymus} sp. (Family: Torymidae).}
\end{figure}

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from eight different plant genotypes that spanned the range of trait
variation observed in this willow population \citep{Barbour2015}. For
the original food web (eight replicates per genotype), we used flagging
tape to mark 14 galled leaves per plant (\textasciitilde{}30 larvae),
allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that all galls had
similar phenology when we collected galls later in the season. For the
removal food web, we enclosed 14 galled leaves with 10x15cm organza bags
(ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species
that attack during larval development. This treatment did not exclude
the egg parasitoid \emph{Platygaster} sp., which attacks prior to gall
initiation \citep[larva initiate gall development in Cecidomyiid
midges:][]{Gagne1989}. It was not possible to reciprocally manipulate
parasitoid attack (i.e.~exclude egg parasitoid, but allow larval
parasitoids), because it was not possible to identify midge oviposition
sites prior to gall formation. In late August, we collected marked and
bagged galls from each plant, placed them into 30 mL vials and kept them
in the lab for 4 months at room temperature. We then opened galls under
a dissecting scope and determined whether larvae survived to pupation
(our measure of fitness) or were parasitized. We did not include other
sources of mortality, such as early larval death, as they could
influence the expression of the gall phenotype and confound estimates of
selection. For the food-web treatment that excluded larval parasitoids,
we further restricted our data by removing any incidental instances of
parasitism by a larval parasitoid. This represented less than 3\% of the
observations in this food-web treatment and allowed us to focus our
inferences of selection on those imposed by the egg parasitoid. Our
final dataset contains survival estimates for 1285 larvae from 613 galls
and 111 plants.

@@

Inferring adaptive landscapes assumes that trait distributions are
multivariate normal \citep{Lande1983}. To approximate this assumption,
we log-transformed clutch size and square-root transformed oviposition
preference. We then scaled all phenotypic traits (mean=0 and SD=1)
across treatments prior to our analyses (detailed below) to ensure that
our estimates of selection were comparable across traits and with other
studies.

Our analysis consisted of four parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
standardized phenotypic traits (\(i\)) of individuals \textemdash in
each food-web treatment. Second, we translated selection surfaces into
the scale of relative fitness (\(w\)) in order to calculate standardized
selection gradients. Third, we used our estimates of directional
selection gradients to characterize the slope of the adaptive landscape,
which we used to measure selective constraints. Finally, we estimated
the curvature of the adaptive landscape and used a simulation to explore
its effects on the adaptive potential of the gall midge population in
the next generation.

\textbf{Selection surfaces}: Since larval survival was our measure of
absolute fitness, we used a GLMM that assumed a binomial error
distribution (and logit-link function). To approximate the selection
surface, we modelled larval survival as a function of food-web treatment
as well as linear (\(\alpha_i\)), quadratic (\(\alpha_{ii}\)), and
linear interactions (\(\alpha_{ij}\)) between each trait. We also
allowed these selection surfaces (\(\alpha\)) to vary between food-web
treatments. Note that to obtain valid estimates of linear selection
surfaces, we removed nonlinear terms prior to estimating linear
relationships \citep{Lande1983}. Other approaches have been advocated
for approximating selection surfaces \citep{Schluter1988}; however, our
approach enables us to calculate selection gradients, and thus is more
appropriate for approximating the adaptive landscape \citep{Arnold2003}.
To account for the nonindependence of clutch size (measured at gall
level) and oviposition preference (measured at plant level) as well as
any independent effects of willow genotype on larval survival, we
modelled gall ID nested within plant ID nested within genotype ID as
random effects. Although statistical models with random effects are not
common in analyses of natural selection, we think that modelling random
effects can mitigate biased estimates of selection due to environmental
covariances between traits and fitness \citep{Rausher1992}. Since our
end goal was to characterize the relationship between mean trait values
and mean fitness (adaptive landscape), we assumed the mean value of our
random effects (i.e.~setting them to zero) when calculating selection
surfaces. We then used parametric bootstrapping (1,000 replicates) to
estimate the effect of food-web treatment on larval survival as well as
selection surfaces in each food-web treatment. To determine whether
trait-fitness relationships differed between food-web treatments, we
calculated the difference in bootstrapped replicates between treatments.

\textbf{Selection gradients}: Selection gradients cannot be estimated
directly from GLMMs of selection surfaces because the response is in
terms of absolute fitness and the coefficients are on a nonlinear scale.
For example, the coefficients in the above model measure the change in
the logarithm of the odds of surviving (i.e.
\(\text{ln}\{W(z)/[1-W(z)]\}\)) associated with 1 SD change in a trait
with all other traits held fixed at their mean. Therefore, we used the
method developed by \citet{Janzen1998} to translate selection surfaces
from the above model into the scale of relative fitness in order to
calculate directional (\(\beta_i\)), quadratic (\(\gamma_{ii}\)), and
correlational (\(\gamma_{ij}\)) selection gradients. Briefly, this
method calculates the average gradient of selection surfaces by
multiplying the average of \(W(z)[1-W(z)]\) by each regression
coefficient (e.g. \(\alpha_i\), \(\alpha_{ii}\), or \(\alpha_{ij}\)). We
then divided this average gradient by the mean fitness (\(\bar W\)) to
put it on the scale of relative fitness, and thus interpretable as a
selection gradient. We estimated selection gradients separately for each
food-web treatment. We also determined whether selection gradients
differed between food-web treatments by calculating the difference in
bootstrapped replicates between treatments. Note that we doubled all
quadratic terms prior to calculating selection gradients to put them on
the same scale as estimates of directional and correlational selection
\citep{Stinchcombe2008}.

\textbf{Selective constraints}: The number of selective constraints is
determined by the slope of the adaptive landscape, which in our study
corresponds to:

\[\textbf{Slope} = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} \]
where each \(\beta_i\) corresponds to the directional selection gradient
acting on each trait. By comparing the number of directional selection
gradients that show clear evidence of contributing to the slope
(i.e.~95\% CI does not overlap zero) in our food-web treatments, we can
infer the effect of food-web structure on selective constraints. This
gives insight as to whether there is a specific combination of
phenotypes favored by natural selection (i.e.~more traits under
directional selection), or whether multiple trait combinations have
comparable fitness (i.e.~less traits under directional selection).

\textbf{Genetic constraints}: The indirect effects of selection on the
G-matrix
(\(\Delta\text{G}=\text{G}(\gamma - \beta \beta^\text{T})\text{G}\)) are
governed by the curvature of the adaptive landscape
(\(\text{C}=\gamma - \beta \beta^\text{T}\)), which in our study
corresponds to:

@@

\[\textbf{Curvature} = \begin{pmatrix} \text{C}_{\text{Diam:Diam}}&& \\ \text{C}_{\text{Clutch:Diam}} & \text{C}_{\text{Clutch:Clutch}} & \\ \text{C}_{\text{Pref:Diam}} & \text{C}_{\text{Pref:Clutch}} & \text{C}_{\text{Pref:Pref}} \end{pmatrix}\]
where each \(\text{C}_{ii}\) (diagonal) represents the effect of
selection on the additive genetic variance in a trait, and each
\(\text{C}_{ij}\) (off-diagonal) represents the effect of selection on
the additive genetic covariance between a particular trait combination.
We used bootstrapped values of each selection gradient to estimate the
curvature of each component of the matrix and its associated 95\% CI. We
also used this information to determine whether the curvature of each
component differed between our food-web treatments.

Knowledge of the curvature matrix alone gives an incomplete picture of
its indirect effect on genetic constraints. This is because genetic
constraints are ultimately determined by the structure of the G-matrix,
and therefore also depend on its structure before selection. Although we
do not know the underlying G-matrix for the traits we measured in this
experiment, we can still gain insight to how our food-web treatment
would alter genetic constraints more generally. Specifically, we
calculated how our best estimate of the curvature matrix (mean values)
in each treatment changed the structure of \ensuremath{10^{4}} random
G-matrices for the next generation. We restricted the range of additive
genetic variance (\emph{V\textsubscript{G}}) in these G-matrices to
between 0 and 0.5 to reflect typical ranges in narrow-sense heritability
values (\emph{h}\textsuperscript{2}; note that
\emph{h}\textsuperscript{2} = \emph{V\textsubscript{G}} when the
phenotypic variance is scaled to 1). Note that this analysis assumes
that the effects of recombination and mutation on the G-matrix are much
smaller than the effects of selection, which appears to be a reasonable
assumption over short time scales \citep{Arnold2008}.

The G-matrix itself is a complex structure, but has a clear theoretical
link to the evolvability of phenotypic traits \citep{Hansen2008}.
Evolvability measures the ability for a trait to evolve toward a given
direction of selection \citep{Hansen2008}. In the absence of knowledge
about the direction of selection that a population will actually
experience in the next generation, we can compute its average
evolvability over random selection gradients
\citep{Hansen2008, Melo2015}. By computing the average evolvability
(here, we used 1000 random \(\beta\)s) for each of \ensuremath{10^{4}}
random G-matrices, we can then assess how changes in the curvature
matrix alter the evolutionary potential of the associated traits. We
report the distribution of these effect sizes, rather than conduct a
statistical test, because the number of replicates in a simulation can
be arbitrarily high, thus making any effect size statistically
significant \citep{White2014}.
